---
title: "glacier_exploratory"
format: html
editor: visual
cache: TRUE
---

## Load packages

```{r}
pacman::p_load('tidyverse',
               'here',
               'janitor',
               'ggthemes',
               'terra',
               'tmap',
               'patchwork',
               'stars',
               'rnaturalearth')
```

## Read data

```{r}
glacier <- st_read(here('data','LIA_glacier_outlines_Alps', 'LIA_Alps_Reinthaler_Paul_2024.shp'))
```

```{r}
lake_30 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '1930s_glaciallakes.tab')) %>% 
  clean_names()

lake_70 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '1970s_glaciallakes.tab'))%>% 
  clean_names()

lake_80 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '1980s_glaciallakes.tab'))%>% 
  clean_names()

lake_90 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '1990s_glaciallakes.tab'))%>% 
  clean_names()

lake_06 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '2006_glaciallakes.tab'))%>% 
  clean_names()


lake_12 <- read_delim(here('data','VianiC_2018', 
                           'datasets', '2012_glaciallakes.tab'))%>% 
  clean_names()
```

Join the `lake` datasets into one - all have the same column names so we can `bind_rows()`

```{r}
lake_joined <- bind_rows(list(lake_30, lake_70, lake_80,
                           lake_06, lake_12))

```

I want to explore these a bit more so lets find the unique values

```{r}
print(paste0('There are ', n_distinct(lake_joined$event), ' glacier lakes in the dataset'))


```

That seems a bit high, and if you open up the table youd notice that the `event` column has repeat rows for lakes at different sites. This is good! However for seeing the number of lakes and comparing total size, we should have some way to aggregate the lakes. I will separate!

```{r}
#| warning: FALSE
lake_joined <- lake_joined %>% 
  separate(event, into = c("lake", NA), sep = "_") # this has the added benefit of renaming event to 'lake'
```

What's it now? 

```{r}
print(paste0('There are ', n_distinct(lake_joined$lake), ' glacier lakes in the dataset'))

```

91 is a lot to plot but 
Now I want to examine 4 things graphically:

1.  How do the glacier lake areas change over time
2.  Where are the glacier lakes
3.  Where are the glaciers
4.  Where do the lakes intersect with the glaciers (aka should I drop any)

##1 How do the glacier lake areas change over time


```{r}
# First we aggregate by glacier lake and year and take a sum of the total area 
lake_joined %>% 
  group_by(lake, date_time) %>% 
  summarise(total_area = sum(perimeter_m)) %>% 
  ggplot(aes(x = date_time, y = total_area,
             color = lake)) +
  geom_point()+
  geom_line()+
  theme(legend.position = 'none')
```
That is unintelligable for now, but I can see some variation. 

## 2.  Where are the glacier lakes
This is going to require I make the glacier lakes have shape. To do that i'm going to make points at each lake and add a buffer of the outline of the lakes. 

```{r}
lake_geo <- lake_joined %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform( crs = 3857) %>% 
  st_buffer( dist = lake_joined$perimeter_m) %>% 
  st_transform(crs = 4326)

# now we can plot 


tm_shape(lake_geo)+
  tm_polygons(fill = 'lake')+
tm_basemap()
```


It looks like our lakes are all over, lets plot out the glaciers and see their extent

## 3.  Where are the glaciers

```{r}
tm_shape(glacier)+
  tm_polygons(fill = 'LIA_ID')+
tm_basemap()
```

I want to crop to the shape of the aosta valley as we have a lot of glaciers and lakes there 

```{r}
italy_map <- ne_states(country = "Italy", returnclass = "sf") 

aosta_map <- italy_map %>% 
  select(province = name, region, geometry) %>% 
  filter(region == "Valle d'Aosta")
```

Now lets crop to that region and plot again 

```{r}
sf::sf_use_s2(FALSE)


glacier <- glacier %>% 
  st_transform(crs(aosta_map)) 

glacier_cropped <- st_crop(glacier, st_bbox(aosta_map))

lake_geo <- lake_geo %>% 
  st_transform(crs(aosta_map)) 

lake_cropped <- st_crop(lake_geo, st_bbox(aosta_map))
```

## 4.  Where do the lakes intersect with the glaciers (aka should I drop any)

now lets make a big aosta plot

```{r}
tmap_mode("view")

tm_basemap(c(StreetMap = "OpenStreetMap", TopoMap = "OpenTopoMap"))+
   tm_tiles(c(CartoDB = "CartoDB.PositronOnlyLabels")) +
tm_shape(aosta_map)+
  tm_borders()+
tm_shape(glacier_cropped)+
  tm_polygons(fill = 'LIA_ID')+
tm_shape(lake_cropped)+
  tm_polygons(fill = 'lake')
```

## Questions for fpm 2
1. What have you learned about your data? Have any potentially interesting patterns emerged? Point to specific visualizations that you created as you describe your findings.

I have learned that the extent of my data is much larger than what I thought it would be. I assumed that the glaciers and lakes would naturally all be in Valle D'Aosta, as that is where I have heard of them being. Instead I found that my glacier data spans the entire alps, with my lakes being concentrated in the Valle. However my lakes stretch as far as the bottom of the French alps. My final visualization showed that the area that I expected to look at, in the Northeast of Aosta, only has a couple of lakes.


2. In FPM #1, you outlined some questions that you wanted to answer using these data. Have you made any strides towards answering those questions? If yes, how so? If no, what next steps do you need to take (e.g. I need to create X plot type, I still need to track down Y data, I need to restructure existing data so that you can visualize it in Z ways, etc.)? Have any new questions emerged?

My second question is a YES! I can definitely use glacier lake size to model glacier recession! I have perimeter and it seems to generally get bigger over time. My first question, so far, is a no. I have to do more research into existing data to see if there is mapped historical glaciers. 

3. What challenges do you foresee encountering with your data? These can be data wrangling and / or visualization challenges.

This data is big and long spanning. The issue is to make any sort of meaningful visualization I need to lower the scope, maybe to a couple lakes or a smaller area in general. Furthermore I will have to do some further exploration on the actual lakes to see where they all reside, maybe an animation to show all ~9 of them. 
I want only miage lake and glacier
```{r}
miage <- lake_cropped %>% 
  filter(lake == 'Miage')

glacier_miage <- glacier_cropped %>% 
  filter(LIA_ID == 1106)
```

```{r}
tmap_mode("plot")

tm_basemap(c(StreetMap = "OpenStreetMap", TopoMap = "OpenTopoMap"))+
   tm_tiles(c(CartoDB = "CartoDB.PositronOnlyLabels")) +
tm_shape(aosta_map)+
  tm_borders()+
tm_shape(glacier_miage)+
  tm_polygons(fill = 'LIA_ID')+
tm_shape(miage)+
  tm_polygons(fill = 'date_time')
```


```{r}
glacier_areas <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                        'ghiacciai_aree',
                        'ghiacciai_aree.shp'))

miage_glacier <- glacier_areas %>% 
  filter(nome == 'GLACIER DU MIAGE')

miage_label <- miage_glacier[1,]

glacier_basins <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_bacini',
               'ghiacciai_bacini.shp'))

glacier_centroid <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_centroidi_2005',
               'ghiacciai_centroidi_2005.shp'))

glacier_lakes <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_laghi_epiglaciali',
                        'ghiacciai_laghi_epiglaciali.shp'))

miage <- glacier_lakes %>% 
  filter(grepl("Miage", nome_lago)) %>% 
  mutate(anno_rilie = as.integer(as.character(anno_rilie)))

glacier_lakes_2021 <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_laghi_epiglaciali_2021',
               'ghiacciai_laghi_epiglaciali_2021.shp'))

miage_2021 <- glacier_lakes_2021 %>% 
  filter(grepl("MIAGE", nome)) %>% 
  mutate(anno_rilie = 2021)


glacial_width <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_larghezze',
               'ghiacciai_larghezze.shp'))

miage_width <- glacial_width %>% 
  filter(grepl("Miage", nome_gs))

glacier_length <-  st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_lunghezze',
               'ghiacciai_lunghezze.shp'))

miage_lenth <-  glacier_length %>% 
  filter(grepl("Miage", nome_gs))

glacier_permimeter <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'ghiacciai_perimetri',
               'ghiacciai_perimetri.shp'))

miage_perimeter <- glacier_permimeter %>% 
  filter(grepl("MIAGE", nome_gs)) %>% 
  mutate(anno = as.integer(as.character(anno)))

idrografia_demanio <- st_read(here::here('data',
                        'u1412_02-19-2026_20-16-06',
                  'idrografia_demanio',
               'idrografia_demanio.shp'))
```



```{r}
tm_basemap(c(StreetMap = "OpenStreetMap", TopoMap = "OpenTopoMap"))+
   tm_tiles(c(CartoDB = "CartoDB.PositronOnlyLabels")) +
tm_shape(miage)+
  tm_polygons(fill = 'blue',
              fill_alpha = .5,
              fill.legend = tm_legend(title = 'Miage Lake 2005'))+
  tm_shape(miage_glacier)+
  tm_polygons(fill = 'white')+
tm_shape(miage_2021)+
  tm_polygons(fill = 'anno_rilie',
              fill_alpha = .5)+
  tm_add_legend(
  labels = c("2005", "2021"), 
  fill = c("blue", "red"), 
  title = "Year"
)
```

```{r}
tm_layout(bg.color = '#021526',
          outer.bg.color = '#002340',
          frame.color = 'white',
          frame.lwd = 2)+
tm_shape(miage) +
  tm_polygons(fill = 'anno_rilie',
              fill.scale = tm_scale_categorical(values = c('lightblue', 'darkblue')),
               fill.legend = tm_legend(show = FALSE))+
   tm_tiles(c(CartoDB = "CartoDB.PositronOnlyLabels"))+
  tm_shape(miage_glacier)+
  tm_polygons(fill = 'white',
              col = '#08316B',
              lwd = 3)+
  tm_shape(miage_2021)+
  tm_polygons(fill = 'anno_rilie',
              fill.scale = tm_scale_categorical(
                values = 'red'),
              fill_alpha = .5,
              fill.legend = tm_legend(show = FALSE))+
tm_add_legend(type = 'polygons', 
              labels = c('1999', '2006', '2021'), 
              fill = c('lightblue', 'darkblue', 'red'),
               bg.color = "#002340",
              text.color = 'white',
              title.color = 'white',
              frame.color = 'white',
              frame.lwd = 2,
              title = 'Year') 

  
```

```{r}

tm_style("cobalt") +
tm_basemap('CartoDB.DarkMatter') +
  tm_tiles(c(CartoDB = "CartoDB.PositronOnlyLabels"))+
tm_shape(miage_glacier)+
  tm_polygons(fill = 'white')+
tm_shape(miage_perimeter)+
 tm_lines(col = 'anno', 
          col.scale = tm_scale_ordinal(values = 'blues'),
          col.legend = tm_legend(title = 'Year'),
         lwd = 4)+
tm_shape(miage_label)+
  tm_text(
    text = 'nome',
    col = 'black',
    xmod = 0, 
    ymod = -3.1,
   angle = -45,
   size  = .7
  )
  
```
```{r}
miage_area <- lake_joined %>% 
  filter(grepl("Miage", lake)) %>% 
  group_by(date_time) %>% 
  summarize(total_area = sum(area_m_2, na.rm = TRUE)) %>% 
  ggplot(aes(x = date_time, y = total_area)) +
  geom_point(color = 'white')+
  geom_line(color = 'white')+
  labs(
    title = 'Miage Lake Area Over Time',
    x = 'Year',
    y = 'Total Area (Square Meters)'
  )+
  theme_tufte()+
  theme(legend.position = 'none',
        plot.background = element_rect(fill = '#002340'),
        axis.title = element_text(color = 'white'),
        axis.text = element_text(color = 'white'),
        plot.title = element_text(color = 'white')
        )

miage_area
```


